A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-10-28 doc/types-pldi2013/intro.tex
More edits to the intro for pldi

In the past decade machine architectures, particularly at the high
performance end of the spectrum, have undergone a revolution.  The
latest supercomputers are now composed of heterogeneous processors
and deep memory hierarchies.  Programming this class of machines is
becoming increasingly difficult.  Current programming systems for these
machines have intricate features for describing parallelism.  However,
these systems relegate to the programmer the responsibility for:
\begin{itemize}
\item the correctness of their parallel declarations
\item the movement and placement of data in the memory hierarchy
\end{itemize}
Coupled with the complexity of these machines, this places a heavy
burden on the programmer.  The reason that the programmer must bear
this load is because current programming systems have few or no
insights into properties of program data (e.g. locality and independence).  
To enable more efficient programming of this class of machines we 
need programming systems capable of reasoning both statically and dynamically
about the structure of program data.

Researchers have previously explored language designs with hierarchical 
decomposition of data to express locality and independence for high 
performance programming. Two recent examples, Sequoia \cite{Fatahalian06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
mechanism to partition the heap recursively into a tree of collections 
of data.  The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that must be fully checked statically 
(see Section~\ref{sec:related} for more discussion of related work).

Our own experience with writing high-performance applications in both 
Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that for many practical situations a
single, static partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions need to be dynamically computed.
Furthermore, it is useful to allow multiple, simultaneous partitions of
the same data, providing different views on to that data.

To support reasoning about the structure
of data in applications with dynamically instantiated data structures, 
we present a type system and operational semantics for Legion\cite{Legion12}.  
Legion is a programming model that presents {\em logical regions} as an
abstraction for capturing {\em locality} (data that will be used together,
and therefore should be colocated) and {\em independence} (disjoint data
that may be operated on in parallel).  Logical regions have the following
properties that enable programmers to dynamically communicate information
about the structure of program data to Legion:
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;

@@

\item  a logical region may be dynamically partitioned in multiple different ways.
\end{itemize}

While data in Legion is organized into a hierarchy of logical region forests, 
computation is organized into a hierarchy of tasks.  Logical regions and tasks interact
through a system of privileges specifying operations that a task
can perform on a logical region argument (read, write, or reduce).  
Tasks are only permitted to access logical regions for which they 
possess privileges.  
%Privileges are passed to a task by their calling task
%or by creating new logical regions.  
%The Legion programming model 
%mandates a functional restriction: a sub-task can only be passed a subset
%of the privileges possesed by its parent task.  As we describe below,
%our type system relies on a combination of static and dynamic checks to 
%enforce this restriction owing to the first-class nature of logical regions.

Legion's ability to describe dynamically instantiated data structures with
multiple dynamic partitions of data does have an important consequence:
aliasing.  Through multiple partitions
of a logical region, a given datum may belong to multiple different regions.  
For example, if a region $R$ is disjointly partitioned three distinct ways 
and each partition assigns every element in $R$ to a
subregion (partitions need not be total in Legion), then every element
in $R$ is included in three different subregions.  This is why we use
the term {\em logical} regions: language-level regions serve to name
sets of data but do not imply anything about physical layout or placement
in the memory hierarchy.\footnote{There
is a separate system of {\em physical regions} at runtime that hold
copies of the data in the language-level logical regions.  Details can
be found in \cite{Legion12}.}
%In fact,
%the Legion implementation routinely maintains multiple distinct
%physical regions, each with a copy of the data in the corresponding
%logical region, for performance reasons \cite{Legion12}.

One consequence of aliasing is that it significantly hinders 
static analyses for proving properties of parallel programs.  
Approaches to dealing with aliasing in the domain
of parallelism have traditionally bifurcated to extremes: Sequoia and DPJ both disallow aliasing
which enables their static analyses at the coarse granularity of regions;  
approaches like transactional memory\cite{} and thread-level speculation\cite{} perform 
fully dynamic analysis at the granularity of individual heap locations.
Alternatively, the type system we present for Legion leverages a hybrid
approach to handling aliasing that enables a combination of static and dynamic
analyses to prove important properties of Legion programs.

The key observation behind our type system is that dynamic alias analysis
is very easy and relatively cheap when done at the granularity of regions 
instead of individual heap locations.  To that end, our type system supports
static verification of all pointer dereferences and passing of region privileges at 
task calls boundaries.  The Legion runtime then leverages its knowledge of logical
region aliasing to perform dynamic safety checks for the unpacking
and packing of logical regions to/from heap with regards to task privileges.  
These checks are then amortized over all the uses of the logical region 
which significantly reduces their cost.

Another important consequence of aliasing is that it often limits 
parallelism.  Legion currently does not attempt to statically prove any two tasks
parallel, but instead relies on dynamic region aliasing checks to discover
parallelism at runtime.  If regions contain many elements, then these checks are
amortized as we will show in Section~\ref{sec:evaluation}.  In addition
to supporting parallel execution in alias-free tasks, Legion also enables
programmer controlled parallel execution in the presence of aliasing.
In many cases, programmers often want to enable parallel execution 
even in the presence of aliased data.  For example, two tasks 
accessing aliased data may only need to be serialized
with respect to each other (e.g. run atomically) or are permitted to run
simultaneously with the programmer using separate synchronization mechanisms
to control access (e.g. locks).  To support these relaxed access modes
Legion provides {\em cohrence modes} on logical regions.  For each region
argument to a task, a programmer specifies the coherence in which other tasks
can access aliased data while the task is running.  Our operational semantics
provides a formal framework with which to reason about the interaction of tasks
accessing aliased in different coherence modes.

We prove the soundess of both the privilges system and the coherence system.
We then use these results to prove a third result: if two {\em siblings} (tasks that
have the same immediate parent task) $t_1$ and $t_2$ are 
{\em non-interfering} (have no ordering requirements for correctness), 
then the any descendant of $t_1$ in the task hierarchy is non-interfering
with any descendant of $t_2$.  This theorem is the basis for correct, distributed
task scheduling in a Legion implementation, which is crucial for
performance; a centralized scheduler would be a bottleneck
because of the large communication latencies in the target
class of machines.

The crucial aspect of this result is that the tasks $t_1$ and $t_2$ only 
need be non-interfering instead of non-aliased.  Thus, through a combination
of static and dynamic analysis, our privilge type system and coherence system
enable us to implement a hierarchical, distributed scheduling algorithm even
in the presence of aliasing introduced by the need for dynamic descriptions of
data via first-class logical regions.  
%All of this is possible only because
%our programming system supports an abstraction for understanding the structure
%of program data.

@@

\item We give experimental evidence for the Legion design.  On three real-world applications, we show that 
dynamic region pointer checks would be expensive, justifying checking this aspect of the type system statically.
We also show that the cost of checking for non-interference of region privileges is low, showing that a much more expressive
and dynamic language with aliasing is not incompatible with high performance (Section~\ref{sec:evaluation}.

>>>>

In the past decade machine architectures, particularly at the high
performance end of the spectrum, have undergone a revolution.  The
latest supercomputers are now composed of heterogeneous processors
and deep memory hierarchies.  Programming this class of machines is
increasingly difficult.  Current programming systems for these
machines have intricate features for describing parallelism, but contain
few abstractions for describing properties of program data
(e.g. independence).  Consequently, the burden of managing
the correctness of parallel computations and movement of data
through the memory hierarchy is placed on the programmer.

%However,
%these systems relegate to the programmer the responsibility for:
%\begin{itemize}
%\item the correctness of their parallel declarations
%\item the movement and placement of data in the memory hierarchy
%\end{itemize}
%Coupled with the complexity of these machines, this places a heavy
%burden on the programmer.  The reason that the programmer must bear
%this load is because current programming systems have few or no
%insights into properties of program data (e.g. locality and independence).  
%To enable more efficient programming of this class of machines we 
%need programming systems capable of reasoning both statically and dynamically
%about the structure of program data.

To combat this problem, researchers have previously explored programming
systems that are capable of reasoning about the structure of data in
parallel programs.  Traditionally, these attempts have taken two approaches.   
The first approach relies on fully static analyses that
have no runtime overhead, but limited expressivity.  Two recent examples, 
Sequoia \cite{Fatahalian06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
mechanism to statically partition the heap recursively into a tree of collections 
of data.  The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that must be fully checked statically 
(see Section~\ref{sec:related} for more discussion of related work).  The
second approach relies on a dynamic analysis of program data that doesn't
limit expressivity, but incurs runtime overhead.  Examples of this approach 
include transactional memory and thread-level speculation.

These two different approaches illustrate different mechanisms for dealing with
the problem of data aliasing in parallel programs.  The fully static approach outlaws aliasing
by requiring a single static partitioning of data.  Our own experience with writing high-performance 
applications in both Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that for many practical situations a
single, static partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions need to be dynamically computed.
Furthermore, it is useful to allow multiple, simultaneous partitions of
the same data, providing different views on to that data.  Alternatively, 
the fully dynamic approach enables arbitrary data aliasing, but requires 
centralized control logic to perform conflict detection between every pair 
of parallel computations. This results in runtime overhead and is 
not scalable on distributed-memory machines.

In this work, we present a type system and operational semantics for
Legion\cite{Legion12}.  Legion is a programming model that supports multiple,
dynamic partitions of program data which can result in aliasing.  We prove
the soundness of our type system and show that through a combination of
static and dynamic analysis we can implement it in a way that minimizes dynamic
runtime checks and results in low actual runtime overhead.  We then show 
that the soundness of our type system enables a scalable, hierarchical 
distributed scheduling algorithm that enables Legion programs to achieve
high performance on current supercomputers.

Legion presents {\em logical regions} as an
abstraction for capturing {\em locality} (data that will be used together,
and therefore should be colocated) and {\em independence} (disjoint data
that may be operated on in parallel).  Logical regions have the following
properties that enable programmers to dynamically communicate information
about the structure of program data to Legion:
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;

@@

\item  a logical region may be dynamically partitioned in multiple different ways.
\end{itemize}
Through multiple partitions
of a logical region, a given datum may belong to multiple different regions.  
For example, if a region $R$ is disjointly partitioned three distinct ways 
and each partition assigns every element in $R$ to a
subregion (partitions need not be total in Legion), then every element
in $R$ is included in three different subregions.  This is why we use
the term {\em logical} regions: language-level regions serve to name
sets of data but do not imply anything about physical layout or placement
in the memory hierarchy.\footnote{There
is a separate system of {\em physical regions} at runtime that hold
copies of the data in the language-level logical regions.  Details can
be found in \cite{Legion12}.}

While data in Legion is organized into a hierarchy of logical region forests, 
computation is organized into a hierarchy of tasks.  Logical regions and tasks interact
through a system of privileges specifying operations that a task
can perform on a logical region argument (read, write, or reduce).  
Tasks are only permitted to access logical regions for which they 
possess privileges.  Privileges are given to a task by their calling task
or by creating new logical regions.  The Legion programming model 
mandates a functional restriction: a sub-task can only be passed a subset
of the privileges possesed by its parent task.

To enforce this restriction, our type system relies on a combination of static 
and dynamic analyses.  The key observation behind our type system is that dynamic
analysis is very easy and relatively cheap when done at the granularity of
regions instead of individual heap locations.  To that end, our type system supports
static verification of all pointer dereferences and passing of region privileges at 
task calls boundaries.  The Legion runtime then leverages its knowledge of logical
region aliasing to perform dynamic safety checks for the unpacking
and packing of logical regions to/from heap with regards to task privileges.  
These checks are then amortized over all the uses of the logical region 
which significantly reduces their cost.

To enable parallelism, Legion again relies on a combination of static and dynamic
analysis to prove that two tasks operate on independent sets of data.  If two tasks
cannot be proven independent statically, the Legion runtime performs dynamic dependence
tests to detect region aliasing.  Since these tests are performed at the granularity of
regions instead of individual heap locations, they are significantly less costly.
In addition to supporting parallel execution for independent tasks, Legion also enables
programmer specified parallel execution in the presence of aliasing.  For example, two tasks 
accessing aliased data may only need to be serialized
with respect to each other (e.g. run atomically) or are permitted to run
simultaneously with the programmer using separate synchronization mechanisms
to control access (e.g. locks).  To support these relaxed access modes
Legion provides {\em cohrence modes} on logical regions.  For each region
argument to a task, a programmer specifies the coherence in which other tasks
can access aliased data while the task is running.  We present an operational semantics
for coherence modes and show that it is sound.

Using our proofs of the soundess of the privilegs and coherence systems, we then 
prove a third result: if two {\em siblings} (tasks that
have the same immediate parent task) $t_1$ and $t_2$ are 
{\em non-interfering} (have no ordering requirements for correctness), 
then the any descendant of $t_1$ in the task hierarchy is non-interfering
with any descendant of $t_2$.  This theorem is the basis for correct, distributed
task scheduling in a Legion implementation, which is crucial for
performance; a centralized scheduler would be a bottleneck
because of the large communication latencies in the target
class of machines.

%The crucial aspect of this result is that the tasks $t_1$ and $t_2$ only 
%need be non-interfering instead of non-aliased.  Thus, through a combination
%of static and dynamic analysis, our privilge type system and coherence system
%enable us to implement a hierarchical, distributed scheduling algorithm even
%in the presence of aliasing introduced by the need for dynamic descriptions of
%data via first-class logical regions.  
%All of this is possible only because
%our programming system supports an abstraction for understanding the structure
%of program data.

@@

\item We give experimental evidence for the Legion design.  On three real-world applications, we show that 
dynamic region pointer checks would be expensive, justifying checking this aspect of the type system statically.
We also show that the cost of checking for non-interference of region privileges is low, showing that a much more expressive
and dynamic language with aliasing is not incompatible with high performance (Section~\ref{sec:evaluation}).
